Stability of variational quantum dynamics with Gaussian approximants

### 2013-10-31 paper.tex

Even more general matrices are sometimes useful.  For instance, the operator
$$M=\pmatrix{0&〈A|\cr |A〉&0}$$
can be applied to columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}^{\rm T}$, giving objects of the same form.  The top left zero is an operator on Hilbert space, while the bottom right one is an $R×R$ matrix.  The general rule is that the things in each row must have the same column dimension, and visa-versa.  Things that have finite dimensional rows and columns will be written as matrices, and things that are infinite in both directions as Hilbert space operators.  These are both just written as letters.

@@

In the notation defined above, Schrödinger's equation is
$$i\hbar |DΨ(t)〉=H|Ψ(t)〉.$$
Here, $|Ψ〉$ denotes a ket-valued function of time, and $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.  The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$.  Therefore it is written outside the ket.  This could be abbreviated to
$$i\hbar |DΨ〉=H|Ψ〉.$$

This is a very simple equation.  The challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand it over a grid grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes merely thousands of particles.  In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
$$|Ψ(t)〉=|ψ(z(t))〉,$$
or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\left(z₁(t),…,z_R(t)\right)$ is a vector of complex parameters.  This imposes two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.

Some methods of computing dynamics, such as those involving single-particle Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.  The above model does not apply to these.  There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which can compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store all the parameters at once.

@@

If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
$$i\hbar|Dψ(z(t))〉Dz(t)=H|ψ(z(t))〉.$$
On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.  This is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to have components outside this subspace.  This is a consequence of the restricted form of the approximation.  We can, however, look for a vector $\dot z(z)$ that minimises the residual
$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\tag\residual$$
and solve the ordinary differential equation
$${dz\over dt}=\dot z(z),\tag\ode$$
giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.  When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.

The Dirac-Frenkel solution is only optimised locally.  There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkey solution $|ψ(w(t))〉$ does.  This would require that, at early times, $i\hbar|Dψ(z(t))〉$ is closer to $H|ψ(z(t))〉$ than $i\hbar|Dψ(w(t))〉$ is to $H|ψ(w(t))〉$, but that $z(t)$ loses the ability to track the dynamics at later times.  When we discuss regularisation below, we will see that this is in fact the usual case.  Global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.

The kind of ansatz considered in this paper has the form 
$$|ψ(z₁,…,z_R)〉=|ψ'(z₁)〉+⋯+|ψ'(z_R)〉.$$
In fact, this form is very general: the DMRG approximants are the only ones I know of that are not a sum of similar terms.  Here, $z_r$ may be a vector of complex numbers, perhaps a weight and the amplitudes of several modes.  In this case, the derivative has the form
$$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&…&|Dψ'(z_R)〉}.$$

I will refer to the kets $|ψ'(z_r)〉$ as the components of $|ψ(z)〉$.  It is interesting to consider the case where these lie in orthogonal subspaces.  This will very nearly be the case, for instance, if the components are coherent states, with sparsely distributed amplitudes.  In this case, the component derivatives $|Dψ'(z_r)〉$ also lie in the orthogonal subspaces, and the total derivative has a block diagonal structure,
$$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&&\cr&\ddots&\cr&&|Dψ'(z_R)〉},$$
where the empty elements in row $r$ are null kets in the subspace of the component $|ψ'(z_r)〉$.
The residual ket thus becomes
$$i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉=
	\pmatrix{|Dψ'(z₁)〉-H|ψ'(z₁)〉\cr\vdots\cr |Dψ'(z_R)〉-H|ψ'(z_R)〉},$$
where the rows are mutually orthogonal.  Therefore, the least squares problem can be solved separately for each component, independently minimising 
$$\left‖|Dψ'(z_r)〉-H|ψ'(z_r)〉\right‖².$$
If the system is a BEC, and the components are coherent states, this problem is equivalent solving the Gross-Pitaevskii Equation.  Heller and others have investigated it for general systems with coherent state components, and found that the amplitudes obey something very like classical dynamics.  Therefore variational dynamics with sparse components is a waste of time: its results should be very similar to truncated Wigner, and, if they differ, there won't be much reason to believe one rather than the other.

@@

@@

The approximant studied in this paper has the form
$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
	=|A〉d,\tag\Ad$$
where $A=\pmatrix{α₁&⋯&α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.  In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.  The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion operator $|A〉$.  Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.  This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.

>>>>

Even more general matrices are sometimes useful.  For instance, the operator
$$M=\pmatrix{0&|A〉\cr 〈A|&0}$$
can be applied to columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}^{\rm T}$, giving objects of the same form.  The top left zero is an operator on Hilbert space, while the bottom right one is an $R×R$ matrix.  The general rule is that the things in each row must have the same column dimension, and visa-versa.  Things that have finite dimensional rows and columns will be written as matrices, and things that are infinite in both directions as Hilbert space operators.  These are both just written as letters.

@@

In the notation defined above, Schrödinger's equation is
$$i\hbar |DΨ(t)〉=H|Ψ(t)〉.$$
Here, $|Ψ〉$ denotes a ket-valued function of time, and $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.  The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$, and is written outside the ket.  Schrödinger's equation could be abbreviated to
$$i\hbar |DΨ〉=H|Ψ〉,$$
equating functions instead of their values.

Schrödinger's equation is very simple.  The main challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand this over a regular basis grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes mere thousands of particles.  In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
$$|Ψ(t)〉=|ψ(z(t))〉,$$
or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\pmatrix{z₁(t)&…&z_R(t)}^{\rm T}$ is a vector of complex parameters.  This poses two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.

Some methods of computing dynamics, such as those involving single-particle Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.  The above model does not apply to these.  There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store them all at once.

@@

If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
$$i\hbar|Dψ(z)〉Dz=H|ψ(z)〉.$$
On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.  This equation is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to lie outside this subspace.  In other words, the restricted form of the approximation does not include the solution to Schrödinger's equation.  We can, however, look for a least-squares solution, a vector $\dot z(z)$ that minimises the residual
$$\bigl\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\bigr\|²,\tag\residual$$
and solve the ordinary differential equation
$$Dz(t)=\dot z(z(t)),\tag\ode$$
giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.  When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.

The Dirac-Frenkel solution is only optimised locally.  There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkel solution $|ψ(w(t))〉$ does.  This would require that, at early times, $i\hbar|Dψ(z(t))〉$ is closer to $H|ψ(z(t))〉$ than $i\hbar|Dψ(w(t))〉$ is to $H|ψ(w(t))〉$, but that $z(t)$ loses the ability to track the dynamics at later times.  When we discuss regularisation below, we will see that this is in fact the usual case.  Global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.

The kind of ansatz considered in this paper has the form 
$$|ψ(z₁,…,z_R)〉=|ψ'(z₁)〉+⋯+|ψ'(z_R)〉.$$
Here, $z_r$ may be a vector of complex numbers, perhaps a weight and the amplitudes of several modes.  In fact, this form is very general: the DMRG approximants are the only ones I know of that are not a sum of similar terms.  In this case, the derivative has the form
$$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&…&|Dψ'(z_R)〉}.$$

I will refer to the kets $|ψ'(z_r)〉$ as the components of $|ψ(z)〉$.  It is interesting to consider the case where these lie in orthogonal subspaces.  This will very nearly be the case, for instance, if the components are coherent states, with sparsely distributed amplitudes.  In this case, the component derivatives $|Dψ'(z_r)〉$ also lie in the orthogonal subspaces, and the total derivative has a block diagonal structure,
$$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&&\cr&\ddots&\cr&&|Dψ'(z_R)〉},$$
where the empty elements in row $r$ are null kets in the subspace of the component $|ψ'(z_r)〉$.  The ket from Equation~\residual\ thus becomes
$$i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉=
	\pmatrix{|Dψ'(z₁)〉{\dot z}₁(z)-H|ψ'(z₁)〉\cr\vdots\cr |Dψ'(z_R)〉{\dot z}_R(z)-H|ψ'(z_R)〉},$$
where the rows are mutually orthogonal.  Therefore, the least squares problem can be solved separately for each component, independently minimising 
$$\bigl‖|Dψ'(z_r)〉{\dot z}_r(z)-H|ψ'(z_r)〉\bigr‖².$$
If the system is a BEC, and the components $|ψ'(z_r)〉$ are coherent states with amplitudes $z_r$, this problem reduces to the Gross-Pitaevskii Equation.  Heller and others have investigated the problem for general systems with coherent state components, and found that the amplitudes obey something very like classical dynamics.  Therefore variational dynamics with sparse components is a waste of time: its results should be similar to those of the truncated Wigner method, and, if they differ, there won't be much reason to believe one rather than the other.

@@

\beginsection{Modifications to Dirac-Frenkel}

@@

The approximant studied in this paper has the form
$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
	=|A〉d,\tag\Ad$$
where $A=\pmatrix{α₁&⋯&α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.  The form on the left is more complicated than the $|A〉d$ one, but it is an analytic function of the parameters, which is convenient for derivatives and optimisation.  Wavelet theorists have studyed the conditions on the set of ampltiudes $A$ that allow any state to be expanded in this form, and the stability of expansions as the density of ampltiudes decreases towards the minimum for completeness.  This will be discussed later.

In the limit of large $R$, this sum approximates an integral over the amplitude $α$, and any state can be approximated arbitrarily well by this form.  The continuous phase-space expansion is the topic of this and the next section.  Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.  This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.
